Pre working group data exploration: Mainly wrt coral cover

### 2024-07-30 julianna_data_exploration.Rmd
Made graph Nic suggested that is kind of synchrony vibes

@@

## Wrangling

```{r}
megaQuadDF
```

>>>>

## Total cover (not change)

@@

### Wrangling

Need to pivot so species are a column

```{r summarize and pivot}
startAlg <- grep("acanthophora_spicifera", colnames(algae))
endAlg <- grep("valonia_ventricosa", colnames(algae))

algae %>% 
  pivot_longer(cols = colnames(algae)[startAlg:endAlg],
               names_to = "Taxa",
               values_to = "Perc_cov") %>% 
  group_by(year, site, island_side, habitat, Taxa) %>% 
  summarize(Mean_cov = mean(Perc_cov),
            Sd_cov = sd(Perc_cov)) -> taxa_long

Check for most abundant things

```{r calculate most abundant throughout time}
taxa_long %>% 
  group_by(habitat, Taxa) %>% 
  summarize(Mean_cov = mean(Mean_cov)) %>% view()
```

Plot all

```{r plot}
# pdf("abundance_time_noTurf.pdf", height = 10, width = 15)

taxa_long %>% 
  filter(Taxa != "algal_turf") %>% 
  ggplot(aes(x = year, y = Mean_cov, group = Taxa, color = Taxa)) +
  geom_line() +
  facet_grid(habitat ~ site) +
  theme_bw() +
  theme(legend.position = "top")

Plot some interesting ones (skipping CCA/peysonnellia/turf; focusing on most abundant things)

```{r plot most abundant}
taxa_long %>% 
  filter(Taxa == "damselfish_turf" |
           Taxa == "lobophora_variegata" |
           Taxa == "turbinaria_ornata" |
           Taxa == "dictyota_bartayresiana" |
           Taxa == "halimeda_minima" |
           Taxa == "asparagopsis_taxiformis" |
           Taxa == "amansia_rhodantha" |
           Taxa == "cyanophyta" |
           Taxa == "halimeda_discoidea" |
           Taxa == "halimeda_opuntia" |
           Taxa == "sargassum_pacificum" |
           Taxa == "halimeda_taenicola" |
           Taxa == "halimeda_sp" |
           Taxa == "halimeda_distorta" 
           ) %>% 
  ggplot(aes(x = year, y = Mean_cov, group = Taxa, color = Taxa)) +
  geom_line() +
  facet_grid(site ~ habitat)

## Change each time (synchrony vibes)

### Wrangle

```{r calculate change per taxa per year}
algae %>% 
  pivot_longer(cols = colnames(algae)[startAlg:endAlg],
               names_to = "Taxa",
               values_to = "Perc_cov") %>% 
  pivot_wider(names_from = year,
              values_from = Perc_cov) %>% 
  mutate(AlgChFrom2007 = `2008` - `2007`,
       AlgChFrom2008 = `2009` - `2008`,
       AlgChFrom2009 = `2010` - `2009`,
       AlgChFrom2010 = `2011` - `2010`,
       AlgChFrom2011 = `2012` - `2011`,
       AlgChFrom2012 = `2013` - `2012`,
       AlgChFrom2013 = `2014` - `2013`,
       AlgChFrom2014 = `2015` - `2014`,
       AlgChFrom2015 = `2016` - `2015`,
       AlgChFrom2016 = `2017` - `2016`,
       AlgChFrom2017 = `2018` - `2017`,
       AlgChFrom2018 = `2019` - `2018`, 
       # skip 2020
       AlgChFrom2021 = `2022` - `2021`, 
       AlgChFrom2022 = `2023` - `2022`) -> algChTaxa
```

```{r get longer with values per taxa}
startChAT <- grep("AlgChFrom2007", colnames(algChTaxa))
endtChAT <- grep("AlgChFrom2022", colnames(algChTaxa))

algChTaxa %>% 
  select(c(site, habitat, site_habitat, transect, quadrat, island_side, Taxa, colnames(algChTaxa)[startChAT:endtChAT])) %>% 
  pivot_longer(cols = colnames(algChTaxa)[startChAT:endtChAT],
               names_to = "Change_year",
               values_to = "Change_in_percent"
               ) %>% 
  # join with the correct year
  mutate(year = case_when(
    Change_year == "AlgChFrom2007" ~ 2007,
    Change_year == "AlgChFrom2008" ~ 2008,
    Change_year == "AlgChFrom2009" ~ 2009,
    Change_year == "AlgChFrom2010" ~ 2010,
    Change_year == "AlgChFrom2011" ~ 2011,
    Change_year == "AlgChFrom2012" ~ 2012,
    Change_year == "AlgChFrom2013" ~ 2013,
    Change_year == "AlgChFrom2014" ~ 2014,
    Change_year == "AlgChFrom2015" ~ 2015,
    Change_year == "AlgChFrom2016" ~ 2016,
    Change_year == "AlgChFrom2017" ~ 2017,
    Change_year == "AlgChFrom2018" ~ 2018,
    Change_year == "AlgChFrom2019" ~ 2019,
    Change_year == "AlgChFrom2021" ~ 2021,
    Change_year == "AlgChFrom2022" ~ 2022,
    TRUE ~ -9999)) -> longChTaxa
```

```{r get averages per site/hab}
longChTaxa %>% 
  group_by(site, habitat, year, island_side, Taxa) %>% 
  summarize(Mean_ch_cov = mean(Change_in_percent, na.rm=TRUE),
            Sd_ch_cov = sd(Change_in_percent, na.rm=TRUE)) %>% 
  # if you take the mean of NAs ONLY, get NaNs--change those to NA's:
  mutate(Mean_ch_cov = case_when(is.nan(Mean_ch_cov) ~ NA,
                                 TRUE ~ Mean_ch_cov)) %>% 
  mutate(habitat = factor(habitat, levels = c("fringing", "backreef", "outer_10", "outer_17"))) -> meanChTaxSiteHab
```

Not sure how to calculate N because **THERE ARE NA'S**!!! I think these are quadrats that are missed

```{r}
# pdf("NicsPlot.pdf", height = 13, width = 10)

meanChTaxSiteHab %>%
  ggplot(aes(x = year, y = Mean_ch_cov, group = Taxa, color = Taxa)) +
  geom_line() +
  theme_bw() +
  facet_grid(site~habitat) +
  theme(legend.position = "top") +
  ylab("Mean change in cover (%)")

Can we super-impose total change in cover on that?

```{r}
algChYr %>% 
  group_by(site, habitat, year, island_side) %>% 
  summarize(Mean_alg_ChAll = mean(Change_in_percent, na.rm = TRUE)) %>% 
  mutate(Mean_alg_ChAll = case_when(is.nan(Mean_alg_ChAll) ~ NA,
                                    TRUE ~ Mean_alg_ChAll)) %>%
  full_join(meanChTaxSiteHab) -> synchDat

synchDat %>% 
  ggplot(aes(x = year)) +
  geom_line(aes(x = year, y = Mean_ch_cov, group = Taxa, color = Taxa)) +
  theme_bw() +
  geom_line(aes(x = year, y = Mean_alg_ChAll+50), color = "black") +
  facet_grid(site~habitat) +
  theme(legend.position = "top") +
  ylab("Mean change in cover (%)") +
  ylim(-50, 150) # maybe this really needs to be done wrt change in stuff without turf?
```
